---
title: "Extracting Maximum Likelihood Points For The Cores"
output: html_document
---

The point is to use the "core genes" (the ones that are always in the core in accordance with all regression comparisons) to calculate the Maximum Likelihood centers of the cores. The data for this script comes from [here](getting.the.data.separated.by.core.counts.html). Be sure to have the output freshly generated in case you use some ambigous naming in other scripts and could have overriden the files.

```{r load libraries}
library("gbra") #becaus gbra has the function AND it loads plyr, dplyr, data.table, etc.
if(!require("ape")){
  install.packages("ape")
  library("ape")
}
if(!require("stringr")){
  install.packages("stringr")
  library("stringr")
}
if(!require("Rphylip")){
  install.packages("Rphylip")
  library("Rphylip")
}
if(!require("doParallel")){
  install.packages("doParallel")
  library("doParallel")
}
```

```{r quick preparation}
self.dir<- "extracting.mle.for.cores.output"
if(!file.exists(self.dir)){
  dir.create(self.dir)
}
self.dir.folder<- function(file=""){
  return(paste(self.dir, file, sep="/"))
}
```

```{r loading data}
#first of all - legend
legend<-load.legend(legend.file = "/home/alext/Developer/gbra/inst/extradata/legend.csv",sep = "\t")
#this table contains all the names in the "gene.name" column
master.table.core.always.core<-read.table(file="master.table.core.always.core.txt", header = TRUE, sep = "\t")
#read the data itself
bh.data.core<-read.table(file="bh.data.minhit_4.txt", header = TRUE, sep="\t")
#select only core genes from the data
bh.data.core.only<-bh.data.core[as.character(master.table.core.always.core$gene.name),]
#attach genome ids to the end (apparently fast enough to run one thread)
bh.data.core.only$ID_QUERY_GENOME<-sapply(rownames(bh.data.core.only),function(i){
  return(strsplit(i,"X",fixed = TRUE)[[1]][1])
})
#wee need the ID_QUERY_ORF column and ID_QUERY_GENOME columns in front of the dataframe to get the MLEs with a standard function
bh.data.core.only$ID_QUERY_ORF<-rownames(bh.data.core.only)
#placind the ID columns in front of the table
bh.data.core.only <- bh.data.core.only[, c(ncol(bh.data.core.only),(ncol(bh.data.core.only)-1),1:nrow(legend))]
```

Now it is time to calculate an MLE estimate for each genome.

```{r extracting the MLEs}
#Calculate MLEs
genome.mles<-getMLEs(df = bh.data.core.only)
#get the gprimes
genome.gprimes<-extract.gprimes(mle.list = genome.mles)
#format the rownames to have exactly the number of spaces fitch wants to have
rownames(genome.gprimes)<-sapply(rownames(genome.gprimes), function(i){
  rn<-str_pad(i, 12, side = "right",pad = " ")
  return(rn)
})
#distance matrix
genome.mle.dist<-as.matrix(dist(genome.gprimes))
head(genome.mle.dist)
#write out as fitch output
write.fitch(dist.mtx = genome.mle.dist, file=self.dir.folder("core.mle.distance.matrix"))
write.table(x=genome.mle.dist, file=self.dir.folder("genome.mle.dist.txt"), sep = "\t", quote = FALSE)
```

```{r plot}
#assign proper names to the rows
genome.mle.dist.for.plot<-genome.mle.dist
rownames(genome.mle.dist.for.plot)<-sapply(rownames(genome.mle.dist),function(i){return(legend$name[which(legend$id_genomes==str_trim(strsplit(i,"X",fixed=TRUE)[[1]][2]))])})
par(mar=c(0,0,0,20), oma=rep(0,4))
plot(as.dendrogram(hclust(as.dist(genome.mle.dist.for.plot),method = "ward.D2")), horiz=TRUE)
par(mar=rep(0,4), oma=rep(0,4))
plot(as.phylo(hclust(as.dist(genome.mle.dist.for.plot),method = "ward.D2")), type = "fan")
```
-----------

So, at this point, it appeared so, that the the hclust is not ideal for the tree generation, and that is why the futher idea is to generate a tree, where the **"outsider"** genomes will participate as **"separate"** organisms.

```{r load the outsiders }
#load the frequent outsiders
master.table.core.outsider.frequently<-read.table(file="master.table.core.outsider.frequently.txt", header=TRUE, sep="\t")
#load the facultative outsiders
master.table.core.outsider.once<-read.table(file="master.table.core.outsider.once.txt", header=TRUE, sep="\t")
#outsider table
master.table.outsiders<-bh.data.core[unique(as.character(master.table.core.outsider.once$gene.name), as.character(master.table.core.outsider.frequently$gene.name)),]
master.table.outsiders$ID_QUERY_ORF<-rownames(master.table.outsiders)
master.table.outsiders$ID_QUERY_GENOME<-sapply(rownames(master.table.outsiders), function(i){
  return(strsplit(i,"X", fixed=TRUE)[[1]][1])
})
master.table.outsiders<-master.table.outsiders[,c((ncol(master.table.outsiders)-1),ncol(master.table.outsiders),1:nrow(legend))]
#calculate mles
outsider.mles<-getMLEs(df = master.table.outsiders)
#get the gprimes
outsider.gprimes<-extract.gprimes(mle.list = outsider.mles)
head(outsider.gprimes)
#substitute the "X" for "Y" to distinguish from genomes
rownames(outsider.gprimes)<-sapply(rownames(outsider.gprimes),function(i){
  return(paste0("Y",strsplit(i,"X",fixed=TRUE)[[1]][2]))
})
rownames(outsider.gprimes)<-sapply(rownames(outsider.gprimes), function(i){
  rn<-str_pad(i, 12, side = "right",pad = " ")
  return(rn)
})
colnames(outsider.gprimes)<-rownames(outsider.gprimes)
#combine with with genomes
total.mles<-rbind(genome.gprimes,outsider.gprimes)
#distance
total.dist<-as.matrix(dist(total.mles))
#write out as fitch output
write.fitch(dist.mtx = total.dist, file=self.dir.folder("total.dist.txt"))
```

Now, we decided to try the cutoff 3 for an outsider to be considered "outsider".

```{r cutoff 3 on outsider}
#select those that appear as outsiders 3 and more times
master.table.outsiders.3.cutoff <- master.table.core.outsider.frequently %>% filter(outsider>=3)
#select the corresponding data
bh.data.3.cutoff<- bh.data.core[as.character(master.table.outsiders.3.cutoff$gene.name),]
head(bh.data.3.cutoff)
#same stuff - attach the id-columns and sort so that the id-columns appear in the beginning
bh.data.3.cutoff$ID_QUERY_ORF<-rownames(bh.data.3.cutoff)
bh.data.3.cutoff$ID_QUERY_GENOME<-sapply(rownames(bh.data.3.cutoff), function(i){
  return(strsplit(i,"X", fixed=TRUE)[[1]][1])
})
bh.data.3.cutoff <- bh.data.3.cutoff[,c((ncol(bh.data.3.cutoff)-1),ncol(bh.data.3.cutoff),1:nrow(legend))]
head(bh.data.3.cutoff)
#get mles
outsider.3.cutoff<- getMLEs(df = bh.data.3.cutoff)
#extract mle gprimes
outsider.3.cutoff.gprimes<- extract.gprimes(outsider.3.cutoff)
#rename the outsider gprimes to mark them out among the core clusters
rownames(outsider.3.cutoff.gprimes)<-sapply(rownames(outsider.3.cutoff.gprimes),function(i){
  return(paste0("Y",strsplit(i,"X",fixed=TRUE)[[1]][2]))
})
#finally assemble a full table
core.plus.outsider.3.cutoff<-rbind(genome.gprimes, outsider.3.cutoff.gprimes)
head(core.plus.outsider.3.cutoff)
#calculate distance
core.plus.outsider.3.cutoff.dist<- as.matrix(dist(core.plus.outsider.3.cutoff))
#save for fitch
write.fitch(dist.mtx = core.plus.outsider.3.cutoff.dist, file = self.dir.folder("core.plus.outsider.3.cutoff.dist.txt"))
#try out Rfitch here
core.plus.outsider.3.cutoff.fitch<- Rfitch(D = core.plus.outsider.3.cutoff.dist, path = "/usr/local/bin", model="ME", power=2, negative=TRUE,global=FALSE,random.order=FALSE, root=FALSE)
write.tree(phy = core.plus.outsider.3.cutoff.fitch, file = self.dir.folder("core.plus.outsider.3.cutoff.fitch.newick"))
```

At this point we decided to create 43 (the number of 3+ outsiders) trees to see what happens if we substitute a core for its outsiders.
But firs we need a function to rename the "X,Y" stuff back onto names.

```{r renaming function}
rename.tree<-function(tree, legend){
  if(!require("tidyr")){
    install.packages("tidyr")
    library("tidyr")
  }
  tree$tip.label<-sapply(tree$tip.label,function(i){
    if(grepl(x=i, pattern = "X",fixed = TRUE)){
      return(paste0(legend$name[which(legend$id_genomes==extract_numeric(i))],"_genome"))
    }else{
      return(paste0(legend$name[which(legend$id_genomes==extract_numeric(i))],"_OUTSIDER>>>>>>"))
    }
  })
  return(tree)
}
```

Lets try renaming..

```{r renaming}
#now do fitch
rf<- Rfitch(D=total.dist, path = "/usr/local/bin", model="ME", power=2, negative=TRUE, global=FALSE,random.order=FALSE, root=FALSE, quiet=TRUE)
write.tree(phy = rf, file = self.dir.folder("total.dist.newick"))
write.tree(phy = rename.tree(tree = rf,legend = legend), file = self.dir.folder("total.dist.names.newick"))
rf<- Rfitch(D=genome.mle.dist, path = "/usr/local/bin", model="ME", power=2, negative=TRUE, global=FALSE,random.order=FALSE, root=FALSE, quiet=TRUE)
write.tree(phy = rf, file = self.dir.folder("cores.only.dist.newick"))
write.tree(phy = rename.tree(tree = rf,legend = legend), file = self.dir.folder("cores.only.dist.names.newick"))
#and the big one
core.plus.outsider.3.cutoff.fitch.names<-rename.tree(tree = core.plus.outsider.3.cutoff.fitch,legend = legend)
write.tree(phy = core.plus.outsider.3.cutoff.fitch.names, file = self.dir.folder("cores.only.3.cut.dist.names.newick"))
plot(core.plus.outsider.3.cutoff.fitch.names, type="fan")
```

Ok, works, so now to the substitutions..

```{r substitutions for outsiders}
#I have used the cluster previouslu, so in case we want any output to be readable (which anyways it won't be cuz there is not sync on threads),
#we need to delete the log file first
if(!file.exists("cluster.txt")){
  file.remove("cluster.txt")
}
#create a dedicated directory
subdir<-"distance"
if(!file.exists(subdir)){
  dir.create(subdir)
}
cluster <- makeCluster(6,outfile = "cluster.txt")
registerDoParallel(cluster)
clusterExport(cl=cluster, varlist = c("outsider.3.cutoff.gprimes","genome.gprimes","subdir","self.dir.folder","self.dir"))
#the cluster is gonna need a bunch of librarires
clusterEvalQ(cl = cluster,library("Rphylip"))
clusterEvalQ(cl = cluster,library("tidyr"))
clusterEvalQ(cl = cluster,library("ape"))
#be cauthious here, because fitch saves the output to a one single file, called "output" and some other files, 
#thereby we need to create a folder and setwd() into that directory prior to running fitch
fitches<-parLapply(cl = cluster,rownames(outsider.3.cutoff.gprimes),function(i){
  #extract the genome id
  gen.num<-as.character(extract_numeric(i))
  #make a separate directory
  dir.create(gen.num)
  #get current directory 
  c.dir<-getwd()
  #cd into the dir
  setwd(gen.num)
  #get all genomes except for the one that is being substitutes by it's outsider counterpart
  g.gp<- genome.gprimes[!grepl(x=rownames(genome.gprimes),pattern = paste0("[X]",gen.num,"(\\b|\\s)"),perl=TRUE),]
  #get the counterpart itself
  substitute.gprime<- outsider.3.cutoff.gprimes[grepl(x=rownames(outsider.3.cutoff.gprimes),pattern = paste0("[Y]",gen.num,"(\\b|\\s)"),perl=TRUE),]
  #combine
  substituted<-rbind(substitute.gprime,g.gp)
  rownames(substituted)[1]<-paste0("Y",gen.num)
  #calculate the distance
  substituted.distance<-dist(substituted)
  fi<- Rfitch(D=substituted.distance,path = "/usr/local/bin", model="ME", power=2, negative=TRUE, global=FALSE,random.order=FALSE, root=FALSE, quiet=TRUE)
  setwd(c.dir)
  unlink(gen.num,recursive = TRUE)
  #plus save the matrix just in case..
  write.table(x=as.matrix(substituted.distance), file=paste(getwd(), subdir, paste0(gen.num,".dmt"), sep="/"), sep="\t", quote = FALSE)
  fi$id<-gen.num
  return(fi)
})
#now for each plot - renames nodes and save as newick
clusterExport(cl=cluster,c("fitches","legend","rename.tree"))
fitches<-parLapply(cl=cluster,X = fitches, fun = function(i){
  return(rename.tree(tree = i, legend = legend))
})
#save newick
clusterExport(cl=cluster,c("fitches"))
fitches<-parLapply(cl=cluster,X = fitches, fun = function(i){
  write.tree(phy = i, file = self.dir.folder(paste0(i$id,".substitute.tree.newick")))
  return(i)
})
#do not forget to cluse the cluster
stopCluster(cluster)
```

