---
title: "Getting the data separated by core counts with a wide margin"
output: html_document
---

So, here we decided the in accordance with `tp=exp(0.15*log(Rsq))`, 
```{r plot tp-rsq dependence}
Rsq<-seq(0,1,by=0.001)
true.positives<-sapply(Rsq,function(i){
  return(exp(0.15*log(i)))
})
plot(x=Rsq,y=true.positives, type="line", xlab="R squared", ylab="True positive")
abline(h=0.8) #this line marks 0.8 true positive margine
```

##Libraries

```{r load libraries}
  library("gbra")
if(!require("reshape2")){
  install.packages("reshape2")
  library("reshape2")
}
if(!require("scales")){
  install.packages("scales")
  library("scales")
}
if(!require("ape")){
  install.packages("ape")
  library("ape")
}
if(!require("stringr")){
  install.packages("stringr")
  library("stringr")
}
if(!require("Rphylip")){
  install.packages("Rphylip")
  library("Rphylip")
}
if(!require("doParallel")){
  install.packages("doParallel")
  library("doParallel")
}
if(!require("doMC")){
  install.packages("doMC")
  library("doMC")
}
if(!require("foreach")){
  install.packages("foreach")
  library("foreach")
}
if(!require("ggplot2")){
  install.packages("ggplot2")
  library("ggplot2")
}
```

##Load initial data

```{r load initial data}
bh.data <- read.table(file = "bh.data.minhit_4.txt")
legend<-load.legend(legend.file = "/home/alext/Developer/gbra/inst/extradata/legend.csv",sep = "\t")
```

Processing the data to split it into three groups:

* a core: is a gene that has over 0.75 probability of bein in it's onwn domain
* an outsider: is a gene that has less then 0.25 probability of being in another domain
* the rest rest is within the margin and is not considered to be clearly separated

This differs just a little bit from the [time when we just split the data in half with 0.5 margin](getting.the.data.separated.by.core.counts.html). Moreover, this time we will be dealing with Rsquared around 0.25 as can be seen on the plot above: this value corresponds to 0.8 true positive ratio, which is rather tolerable.

```{r read models}
dir<-"data" #data folder
factor.core<-c("core","outsider","nosep") #generate three factors to mark states
if(!file.exists("master.table.core.0.25.0.75.margin.txt")){
  #generate a matrix big enough to hold the data (essentially the same size as the bh.data), initially filled in with "outsider"
  master.table.core.0.25.0.75.margin<-as.data.frame(matrix(nrow=nrow(bh.data),ncol=nrow(legend),factor.core[3]),stringsAsFactors = FALSE)
  #copy the row names from bh.data
  row.names(master.table.core.0.25.0.75.margin)<-row.names(bh.data)
  #copy colnames fromt he legend
  colnames(master.table.core.0.25.0.75.margin)<-legend$id_genomes
  #now, for each row in the initial bh.table go over the models one by one, load them into ram and see where the row gene is placed 
  #relative to each genome
  registerDoMC(12)
  log.file<-"cluster.log"
  if(file.exists(log.file)){
    file.remove(log.file)
  }
  master.table.core.0.25.0.75.margin.list<-foreach(i=1:nrow(legend), .errorhandling = "stop", .verbose = TRUE)%dopar%{
    master.table.core.0.25.0.75.margin.i.j<-foreach(j = i:nrow(legend),.verbose = TRUE)%do%{
      #i==j can not be separated, so just return a nonsep list
      sink(log.file, append=TRUE);cat(paste("Doing", i, j, "pair.\n"));sink()
      #load the corresponding model
      if(j!=i){
        load(file = paste(dir,paste(paste(legend$id_genomes[i],legend$id_genomes[j],sep="_"),".rda",sep = ""),sep = "/"))
      }
      sink(log.file, append=TRUE);cat(paste("Loaded", i, j, "pair.\n"));sink()
      #see if the model R squared score if over 0.75, which makes sence in terms of true positive ratio (needs link)
      if(j==i||as.numeric(l.m.g$Rsq)<0.25){
        genome1<-select.genomes(df = bh.data, g.ids = c(legend$id_genomes[i]))
        genome1.is<-rep("nosep", nrow(genome1))
        names(genome1.is)<-rownames(genome1)
        genome2<-select.genomes(df = bh.data, g.ids = c(legend$id_genomes[j]))
        genome2.js<-rep("nosep", nrow(genome2))
        names(genome2.js)<-rownames(genome2)
        sink(log.file, append=TRUE);cat(paste("Returning artificial for", i, j, "pair.\n"));sink()
        return(list(is=genome1.is, js=genome2.js, i=i, j=j))
      }else{
        #for the pair of genomes - restrict the bh.table to only those selected
        data.tab<-select.genomes(df = bh.data, g.ids = c(legend$id_genomes[i],legend$id_genomes[j]))
        #attempt to predict the probabilities for each pair with a precalculated logistic regression
        glm.probs<-predict(l.m.g$fit, data.tab, type = "response")
        #actually split into three groups with a 0.25, 0.25:0.75, 0.75 probability groups
        glm.probs.groups<-sapply(glm.probs,function(t){
          if(t>=0.75){
            return(FALSE)
          }else if(t<=0.25){
            return(TRUE)
          }else{
            return(NA)
          }
        })
        #append another column, which can be either true of false to indicate the clusters 
        data.tab$cluster<-grepl(pattern = paste("\\b",legend$id_genomes[i],"X",sep = "",collapse = ""),x = rownames(data.tab),perl = TRUE)
        #see if the clustering assignment matches the true clusters, mark as core if yes, otherwise - outsider
        sep.factor.v<-sapply(1:length(glm.probs.groups), function(t){
          if(is.na(glm.probs.groups[t])){
            return(factor.core[3])
          }
          if(glm.probs.groups[t]==data.tab$cluster[t]){
            return(factor.core[1])
          }else{ 
            return(factor.core[2])
          }
        })
        names(sep.factor.v)<-names(glm.probs.groups)
        #assign common name identificators like "X1,2..." in accordance with the legend
        sep.factor.v.j<-sep.factor.v[grepl(pattern = paste("\\b",legend$id_genomes[j],"X",sep = "",collapse = ""),x = names(sep.factor.v),perl = TRUE)]
        sep.factor.v.i<-sep.factor.v[grepl(pattern = paste("\\b",legend$id_genomes[i],"X",sep = "",collapse = ""),x = names(sep.factor.v),perl = TRUE)]
        core.col.i<-which(legend$id_genomes==legend$id_genomes[i])
        core.col.j<-which(legend$id_genomes==legend$id_genomes[j])
        #mark i and j cells (because the matrix here is not triangle and needs symmetry)
        master.table.core.0.25.0.75.margin.i<-foreach(k=1:length(sep.factor.v.i),.combine = "c")%do%{
          return(master.table.core.0.25.0.75.margin[names(sep.factor.v.i)[k],core.col.j]<-sep.factor.v.i[k])
        }
        master.table.core.0.25.0.75.margin.j<-foreach(k=1:length(sep.factor.v.j),.combine = "c")%do%{
          return(master.table.core.0.25.0.75.margin[names(sep.factor.v.j)[k],core.col.i]<-sep.factor.v.j[k])
        }
        sink(log.file, append=TRUE);cat(paste("Returning real for", i, j, "pair.\n"));sink()
        return(list(is=master.table.core.0.25.0.75.margin.i,js=master.table.core.0.25.0.75.margin.j,i=core.col.i,j=core.col.j))
      }
    }
    sink(log.file, append=TRUE);cat(paste("Done full row for", i, "\n"));sink()
    return(master.table.core.0.25.0.75.margin.i.j)
  }
  #save the list cuz it takes hellalot of time to calculate
  save(file = "master.table.core.0.25.0.75.margin.list.rda", master.table.core.0.25.0.75.margin.list)
  #so now to concatinate all i into sum matrices
  master.table.core.0.25.0.75.margin.list.cbind.is<-lapply(master.table.core.0.25.0.75.margin.list, function(x){
    l.is<-do.call(mapply, c(cbind,lapply(x,function(i){
      return(i$is)
    })))
    return(l.is)
  })
  #js are columns in fact
  master.table.core.0.25.0.75.margin.list.cbind.js<-lapply(1:nrow(legend), function(x){
    l.js<-lapply(master.table.core.0.25.0.75.margin.list[[x]],function(j){
      return(j$js)
    })
    return(l.js)
  })
  master.table.core.0.25.0.75.margin.list.cbind.js<-lapply(1:nrow(legend),function(x){
        l.js<-do.call(mapply, c(cbind,lapply(1:(nrow(legend)-length(master.table.core.0.25.0.75.margin.list.cbind.js[[x]])+1),function(j){
          return(master.table.core.0.25.0.75.margin.list.cbind.js[[j]][[x-j+1]])
        })))
        return(l.js)
  })
  #now concatenate all into one matrix
  master.table.core.0.25.0.75.margin.list.cbind<-lapply(1:nrow(legend),function(i){
    dt<-as.data.frame(t(rbind(master.table.core.0.25.0.75.margin.list.cbind.is[[i]],
                                 master.table.core.0.25.0.75.margin.list.cbind.js[[i]])))
    return(dt)
  })
  master.table.core.0.25.0.75.margin<- as.data.frame(rbind_all(master.table.core.0.25.0.75.margin.list.cbind))
  head(master.table.core.0.25.0.75.margin)
  rownames(master.table.core.0.25.0.75.margin)<- unlist(sapply(master.table.core.0.25.0.75.margin.list.cbind,rownames))
  colnames(master.table.core.0.25.0.75.margin)<-sapply(legend$id_genomes,function(i){return(paste0("X",i))})
  #the last column is not needed
  master.table.core.0.25.0.75.margin<-master.table.core.0.25.0.75.margin[,-ncol(master.table.core.0.25.0.75.margin)]
  head(master.table.core.0.25.0.75.margin)
  #finally save out the resulting table
  write.table(x=master.table.core.0.25.0.75.margin, file="master.table.core.0.25.0.75.margin.txt", sep="\t", quote = FALSE)
}
```

Ok, let's suppose for a second that after almost a week of trying to get the matrix right:

```{r get the final results}

#read the data
master.table.core.0.25.0.75.margin<-read.table(file="master.table.core.0.25.0.75.margin.txt")
#create a parallel cluster with the cluster log output
if(file.exists(log.file)){
  file.remove(log.file)
}
cluster <- makeCluster(6,outfile = log.file)
registerDoParallel(cluster)
#create a diriving summary of genes
master.table.core.0.25.0.75.margin.per.gene.summary<-as.data.frame(t(parApply(cl=cluster, X = master.table.core.0.25.0.75.margin, MARGIN = 1, FUN = function(i){
  sum.core<-length(i[i=="core"])
  sum.outsider<-length(i[i=="outsider"])
  sum.nosep<-length(i[i=="nosep"])
  return(c(core=sum.core, outsider=sum.outsider, nonsep=sum.nosep))
})))
#append the names because the following filter will erase the rownames
master.table.core.0.25.0.75.margin.per.gene.summary$gene.name<-row.names(master.table.core.0.25.0.75.margin.per.gene.summary)
#do not forget to cluse the cluster
stopCluster(cluster)
#separate the "always core table"
master.table.core.0.25.0.75.margin.always.core<- master.table.core.0.25.0.75.margin.per.gene.summary %>% filter(outsider==0)
head(master.table.core.0.25.0.75.margin.always.core)
#separate the "outsider once"
master.table.core.0.25.0.75.margin.outsider.once<- master.table.core.0.25.0.75.margin.per.gene.summary %>% filter(outsider==1)
head(master.table.core.0.25.0.75.margin.once)
#separate the frequent outsider
master.table.core.0.25.0.75.margin.outsider.frequently<- master.table.core.0.25.0.75.margin.per.gene.summary %>% filter(outsider>1)
head(master.table.core.0.25.0.75.margin.outsider.frequently)
#write out the data
write.table(master.table.core.0.25.0.75.margin.always.core, 
            file="master.table.core.0.25.0.75.margin.always.core.txt", sep="\t", col.names=TRUE, quote = FALSE)
write.table(master.table.core.0.25.0.75.margin.outsider.once, 
            file="master.table.core.0.25.0.75.margin.outsider.once.txt", sep="\t", col.names=TRUE, quote = FALSE)
write.table(master.table.core.0.25.0.75.margin.outsider.frequently, 
            file="master.table.core.0.25.0.75.margin.outsider.frequently.txt", sep="\t", col.names=TRUE, quote = FALSE)
#
```

At this moment i got some headache and got frustrated, so i just copypaste some working code `extracting.mle.from.cores.Rmd`
I just simply copypasted it below, and changed the filenames.

```{r loading data}
#first of all - legend
legend<-load.legend(legend.file = "/home/alext/Developer/gbra/inst/extradata/legend.csv",sep = "\t")
#this table contains all the names in the "gene.name" column
master.table.core.always.core<-read.table(file="master.table.core.0.25.0.75.margin.always.core.txt", header = TRUE, sep = "\t")
#read the data itself
bh.data.core<-read.table(file="bh.data.minhit_4.txt", header = TRUE, sep="\t")
#select only core genes from the data
bh.data.core.only<-bh.data.core[as.character(master.table.core.always.core$gene.name),]
#attach genome ids to the end (apparently fast enough to run one thread)
bh.data.core.only$ID_QUERY_GENOME<-sapply(rownames(bh.data.core.only),function(i){
  return(strsplit(i,"X",fixed = TRUE)[[1]][1])
})
#wee need the ID_QUERY_ORF column and ID_QUERY_GENOME columns in front of the dataframe to get the MLEs with a standard function
bh.data.core.only$ID_QUERY_ORF<-rownames(bh.data.core.only)
#placind the ID columns in front of the table
bh.data.core.only <- bh.data.core.only[, c(ncol(bh.data.core.only),(ncol(bh.data.core.only)-1),1:nrow(legend))]
```

Now it is time to calculate an MLE estimate for each genome.

```{r extracting the MLEs}
#Calculate MLEs
genome.mles<-getMLEs(df = bh.data.core.only)
#get the gprimes
genome.gprimes<-extract.gprimes(mle.list = genome.mles)
#format the rownames to have exactly the number of spaces fitch wants to have
rownames(genome.gprimes)<-sapply(rownames(genome.gprimes), function(i){
  rn<-str_pad(i, 12, side = "right",pad = " ")
  return(rn)
})
#distance matrix
genome.mle.dist<-as.matrix(dist(genome.gprimes))
head(genome.mle.dist)
#write out as fitch output
write.fitch(dist.mtx = genome.mle.dist, file="core.mle.distance.matrix")
write.table(x=genome.mle.dist, file="genome.mle.dist.txt", sep = "\t", quote = FALSE)
```

```{r plot}
#assign proper names to the rows
genome.mle.dist.for.plot<-genome.mle.dist
rownames(genome.mle.dist.for.plot)<-sapply(rownames(genome.mle.dist),function(i){return(legend$name[which(legend$id_genomes==str_trim(strsplit(i,"X",fixed=TRUE)[[1]][2]))])})
par(mar=c(0,0,0,20), oma=rep(0,4))
plot(as.dendrogram(hclust(as.dist(genome.mle.dist.for.plot),method = "ward.D2")), horiz=TRUE)
par(mar=rep(0,4), oma=rep(0,4))
plot(as.phylo(hclust(as.dist(genome.mle.dist.for.plot),method = "ward.D2")), type = "fan")
```
-----------

So, at this point, it appeared so, that the the hclust is not ideal for the tree generation, and that is why the futher idea is to generate a tree, where the **"outsider"** genomes will participate as **"separate"** organisms.

```{r load the outsiders }
#load the frequent outsiders
master.table.core.outsider.frequently<-read.table(file="master.table.core.0.25.0.75.margin.outsider.frequently.txt", header=TRUE, sep="\t")
#load the facultative outsiders
master.table.core.outsider.once<-read.table(file="master.table.core.0.25.0.75.margin.outsider.once.txt", header=TRUE, sep="\t")
#outsider table
master.table.outsiders<-bh.data.core[unique(as.character(master.table.core.outsider.once$gene.name), as.character(master.table.core.outsider.frequently$gene.name)),]
master.table.outsiders$ID_QUERY_ORF<-rownames(master.table.outsiders)
master.table.outsiders$ID_QUERY_GENOME<-sapply(rownames(master.table.outsiders), function(i){
  return(strsplit(i,"X", fixed=TRUE)[[1]][1])
})
master.table.outsiders<-master.table.outsiders[,c((ncol(master.table.outsiders)-1),ncol(master.table.outsiders),1:nrow(legend))]
#calculate mles
outsider.mles<-getMLEs(df = master.table.outsiders)
#get the gprimes
outsider.gprimes<-extract.gprimes(mle.list = outsider.mles)
head(outsider.gprimes)
#substitute the "X" for "Y" to distinguish from genomes
rownames(outsider.gprimes)<-sapply(rownames(outsider.gprimes),function(i){
  return(paste0("Y",strsplit(i,"X",fixed=TRUE)[[1]][2]))
})
rownames(outsider.gprimes)<-sapply(rownames(outsider.gprimes), function(i){
  rn<-str_pad(i, 12, side = "right",pad = " ")
  return(rn)
})
colnames(outsider.gprimes)<-rownames(outsider.gprimes)
#combine with with genomes
total.mles<-rbind(genome.gprimes,outsider.gprimes)
#distance
total.dist<-as.matrix(dist(total.mles))
#write out as fitch output
write.fitch(dist.mtx = total.dist, file="total.dist.txt")
```
Now, we decided to try the cutoff 3 for an outsider to be considered "outsider".
```{r cutoff 3 on outsider}
#select those that appear as outsiders 3 and more times
master.table.outsiders.3.cutoff <- master.table.core.outsider.frequently %>% filter(outsider>=3)
#select the corresponding data
bh.data.3.cutoff<- bh.data.core[as.character(master.table.outsiders.3.cutoff$gene.name),]
head(bh.data.3.cutoff)
#same stuff - attach the id-columns and sort so that the id-columns appear in the beginning
bh.data.3.cutoff$ID_QUERY_ORF<-rownames(bh.data.3.cutoff)
bh.data.3.cutoff$ID_QUERY_GENOME<-sapply(rownames(bh.data.3.cutoff), function(i){
  return(strsplit(i,"X", fixed=TRUE)[[1]][1])
})
bh.data.3.cutoff <- bh.data.3.cutoff[,c((ncol(bh.data.3.cutoff)-1),ncol(bh.data.3.cutoff),1:nrow(legend))]
head(bh.data.3.cutoff)
#get mles
outsider.3.cutoff<- getMLEs(df = bh.data.3.cutoff)
#extract mle gprimes
outsider.3.cutoff.gprimes<- extract.gprimes(outsider.3.cutoff)
#rename the outsider gprimes to mark them out among the core clusters
rownames(outsider.3.cutoff.gprimes)<-sapply(rownames(outsider.3.cutoff.gprimes),function(i){
  return(paste0("Y",strsplit(i,"X",fixed=TRUE)[[1]][2]))
})
#finally assemble a full table
core.plus.outsider.3.cutoff<-rbind(genome.gprimes, outsider.3.cutoff.gprimes)
head(core.plus.outsider.3.cutoff)
#calculate distance
core.plus.outsider.3.cutoff.dist<- as.matrix(dist(core.plus.outsider.3.cutoff))
#save for fitch
write.fitch(dist.mtx = core.plus.outsider.3.cutoff.dist, file = "core.plus.outsider.3.cutoff.dist.txt")
#try out Rfitch here
core.plus.outsider.3.cutoff.fitch<- Rfitch(D = core.plus.outsider.3.cutoff.dist, path = "/usr/local/bin", model="ME", power=2, negative=FALSE,global=FALSE,random.order=FALSE, root=FALSE)
write.tree(phy = core.plus.outsider.3.cutoff.fitch, file = "core.plus.outsider.3.cutoff.fitch.newick")
core.plus.outsider.3.cutoff.names.fitch<-rename.tree(tree = core.plus.outsider.3.cutoff.fitch, legend = legend)
write.tree(phy = core.plus.outsider.3.cutoff.names.fitch, file = "core.plus.outsider.3.cutoff.fitch.names.newick")
#do fitch for the core genomes
genome.mle.fitch<- Rfitch(D=genome.mle.dist, path = "/usr/local/bin", model="ME",
                          power=2, negative=FALSE, global=FALSE,random.order=FALSE, root=FALSE, quiet=TRUE)
genome.mle.fitch.names<- rename.tree(tree = genome.mle.fitch, legend = legend)
write.tree(phy = genome.mle.fitch.names, file = "genome.mle.fitch.names.newick")
```
At this point we decided to create 43 (the number of 3+ outsiders) trees to see what happens if we substitute a core for its outsiders.
But firs we need a function to rename the "X,Y" stuff back onto names.
```{r renaming function}
rename.tree<-function(tree, legend){
  tree$tip.label<-sapply(tree$tip.label,function(i){
    if(grepl(x=i, pattern = "X",fixed = TRUE)){
      return(paste0(legend$name[which(legend$id_genomes==extract_numeric(i))],"_genome"))
    }else{
      return(paste0(legend$name[which(legend$id_genomes==extract_numeric(i))],"_OUTSIDER>>>>>>"))
    }
  })
  return(tree)
}
```
Lets try renaming..
```{r renaming}
core.plus.outsider.3.cutoff.fitch.names<-rename.tree(tree = core.plus.outsider.3.cutoff.fitch,legend = legend)
write.tree(phy = core.plus.outsider.3.cutoff.fitch.names, file = "core.plus.outsider.3.cutoff.fitch.names.newick")
plot(core.plus.outsider.3.cutoff.fitch.names, type="fan")
```
Ok, works, so now to the substitutions..
```{r substitutions for outsiders}
#I have used the cluster previouslu, so in case we want any output to be readable (which anyways it won't be cuz there is not sync on threads),
#we need to delete the log file first
file.remove("cluster.txt")
#create a dedicated directory
subdir<-"distance"
if(!file.exists(subdir)){
  dir.create(subdir)
}
cluster <- makeCluster(6,outfile = "cluster.txt")
registerDoParallel(cluster)
clusterExport(cl=cluster, varlist = c("outsider.3.cutoff.gprimes","genome.gprimes","subdir"))
#the cluster is gonna need a bunch of librarires
clusterEvalQ(cl = cluster,library("Rphylip"))
clusterEvalQ(cl = cluster,library("tidyr"))
clusterEvalQ(cl = cluster,library("ape"))
#be cauthious here, because fitch saves the output to a one single file, called "output" and some other files, 
#thereby we need to create a folder and setwd() into that directory prior to running fitch
fitches<-parLapply(cl = cluster,rownames(outsider.3.cutoff.gprimes),function(i){
  #extract the genome id
  gen.num<-as.character(extract_numeric(i))
  #make a separate directory
  dir.create(gen.num)
  #get current directory 
  c.dir<-getwd()
  #cd into the dir
  setwd(gen.num)
  #get all genomes except for the one that is being substitutes by it's outsider counterpart
  g.gp<- genome.gprimes[!grepl(x=rownames(genome.gprimes),pattern = paste0("[X]",gen.num,"(\\b|\\s)"),perl=TRUE),]
  #get the counterpart itself
  substitute.gprime<- outsider.3.cutoff.gprimes[grepl(x=rownames(outsider.3.cutoff.gprimes),pattern = paste0("[Y]",gen.num,"(\\b|\\s)"),perl=TRUE),]
  #combine
  substituted<-rbind(substitute.gprime,g.gp)
  rownames(substituted)[1]<-paste0("Y",gen.num)
  #calculate the distance
  substituted.distance<-dist(substituted)
  fi<- Rfitch(D=substituted.distance,path = "/usr/local/bin", model="ME", power=2, negative=FALSE, global=FALSE,random.order=FALSE, root=FALSE, quiet=TRUE)
  setwd(c.dir)
  unlink(gen.num,recursive = TRUE)
  #plus save the matrix just in case..
  write.table(x=as.matrix(substituted.distance), file=paste(getwd(), subdir, paste0(gen.num,".dmt"), sep="/"), sep="\t", quote = FALSE)
  fi$id<-gen.num
  return(fi)
})
#now for each plot - renames nodes and save as newick
clusterExport(cl=cluster,c("fitches","legend","rename.tree"))
fitches<-parLapply(cl=cluster,X = fitches, fun = function(i){
  return(rename.tree(tree = i, legend = legend))
})
#save newick
clusterExport(cl=cluster,c("fitches"))
fitches<-parLapply(cl=cluster,X = fitches, fun = function(i){
  write.tree(phy = i, file = paste0(i$id,".substitute.tree.newick"))
  return(i)
})
#do not forget to cluse the cluster
stopCluster(cluster)
```
